Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
edited through

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must record each individual's genome.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
To make matters worse,
a population of size $N$ must be simulated across many multiples of $N$ generations
to produce stable genetic patterns \citep{wright1931evolution, wakeley2005coalescent}.
Because of this computational burden, even the fastest simulation frameworks such as
SLiM 2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models of large genomic regions without selection~\citep{fwdpp,Hernandez2015-wf},
and existing simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
This is equivalent to knowing the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral meiosis.
The embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much much more than typically recorded in a forward simulation, however,
so further savings can be had by discarding all information irrelevant to the genetic history
of the extant population.

In this paper, we describe a method for storing a \emph{succinct tree sequence} 
(or, for brevity, ``tree sequence''),
and, hence, genome sequence, as well as an algorithm for simplifying these.
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe a general method of using these tools to efficiently record,
and later process, the population pedgiree in forwards-time simulation.
While providing substantial savings in time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

We begin by benchmarking the performance improvement achieved by connecting
a previously published forwards-time simulation library,
\fwdpp{} \citep{fwdpp},
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

To measure the performance gains from recording the pedigree we ran simulations both with and without recording.
All simulations implemented a discrete-time Wright-Fisher population of diploid
population size $N$, simulated for $10N$ generations.
The simulations without pedigree recording introduced neutral mutations at a rate
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations per gamete and recombination breakpoints per
diploid,
respectively, each generation.
% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We ran simulations with different values of $N$ and varied the size of the region according to the scaled recombination
parameter $\rho = 4Nr$.  When simulating neutral mutations, we kept the scaled mutation parameter, $\theta = 4N\mu$
equal to $\rho$.

@@

We implemented our method in \cpp{}, using \fwdpp{} library functions and
interface code (details below), using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{}, a Python package based
on \fwdpp.
% $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
% We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

%% \krt{currently, we are not doing anything with the detailed timings.  I think that is fine, and that we can delete last
%%sentence of preceding paragraph.}

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
Pedigree tracking
dramatically improves run times (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
(Figure~\ref{fig:relative_speedup_selection}).  Pedigree tracking reduces runtimes more for larger
population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when tracking
pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative
behavior with a numerically larger benefit to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

In our test runs, our method used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
        different diploid population sizes ($N$). The left column shows run times for
        simulations including the tracking of neutral mutations. The right column of
panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
the results for an implementation using \fwdpy{} where the pedigree simplification steps were handled in a separate thread of
execution and fitness calculations were parallelized.}
\end{figure}

\begin{figure}
    \includegraphics[]{sims/speedup}
    \caption{\label{fig:relative_speedup_selection}Relative performance improvement due to pedigree tracking. This figure
    shows the ratio of the run time tracking neutral mutations over the run time tracking the pedigree. Data points are taken
from Figure~\ref{fig:runtimes_selection} for simulations that ran to completion using both methods.}
\end{figure}

@@

%%%%%%
\subsection*{Working with a tree sequence through tables}

The simulations above recorded information about each new individual
in a collection of tables that together define a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
A closely related structure is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1991two,griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper.
ARGs are rarely used
in practise. In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime\ coalescent simulator.

Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a much larger class of tree topology.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence,
and each node refers to a distinct ancestor.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with.
Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born one and two units of time in the past, respectively.

@@

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records:
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees in this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding compact and algorithmically efficient.

Recovering the sequence of trees from this information is straightforward:
each point along the genome that the tree topology changes
is accompanied by the end of some {edges} and the beginning of others.
Since each {edge} records the genomic interval
over which a given node inherits from a particular ancestor,
to construct the tree at a certain point in the genome
we need only retrieve all edges overlapping that point
and construct the corresponding tree.
To modify the tree to reflect the genealogy at a nearby location,
we simply remove those edges whose intervals do not overlap that location,
and add those new edges whose intervals do.
Incidentally, this property that edges naturally encode \emph{differences}
between nearby trees (e.g., as ``subtree prune and regraft'' moves)
allows for efficient algorithms to compute statistics of the genome sequence that take advantage
of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site,
and results in a specific derived state.
Thus, all samples below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we ran a simulation of 500,000 samples of a
$200$ megabase human-like chromosome ($N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation) using \msprime. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the nodes, edges, sites and mutations (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB (giving a compression factor of 144,000 in
this instance).
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time just to read the information into memory.

@@

The facilities for working with succinct tree sequences are implemented as part
of the \msprime\ Python API, which provides a powerful platform for processing
tree topology and mutation data. The new portions of \msprime{} that we discuss
here are dedicated to tree sequence input and output using simple tables of
data, so we refer to this as the `Tables API'. The four key tables (nodes,
edges, sites and mutations) are described above.

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must record each individual's genome.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
carrying full genotypes for even modest numbers of individuals through a simulation
can quickly become prohibitive.
To make matters worse,
a population of size $N$ must be simulated across many multiples of $N$ generations
to produce stable genetic patterns \citep{wright1931evolution, wakeley2005coalescent}.
Because of this computational burden, even the fastest simulation frameworks such as
SLiM 2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations.
%% plr: not sure why we need to say these things here?
% and there are still near-order-of-magnitude performance differences between these implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models of large genomic regions without selection~\citep{fwdpp,Hernandez2015-wf},
and existing simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation:
it does not matter if we generate each generation's mutations during the simulation, or afterwards.
Precisely, we need to know the genealogical trees relating all sampled individuals
at each position along the genome.
Combined with ancestral genotypes and the origins of new mutations,
these trees completely specify the genomic sequence of any individual in the population at any time.
To obtain this information, we record from forward simulation the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral meiosis,
periodically discarding all information irrelevant to the genetic history
of the extant population.
The information in this embellished pedigree is stored as a \emph{succinct tree sequence} 
(or, for brevity, ``tree sequence''),
which contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome.

In this paper, we describe a storage method for \emph{succinct tree sequences} 
and, hence, genome sequence, as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information.
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

The strategy described above is only of interest if it is computationally feasible.
Therefore,
we begin by benchmarking the performance improvement achieved by this method,
implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
and tree sequence tools implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences succinctly in several \emph{tables};
(b) an algorithm to record these tables on the fly during a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

To measure the performance gains from recording the pedigree we ran simulations 
both with and without recording.
(Although we record more than just the parent--offspring relationships of the pedigree,
for brevity we refer to the method as ``pedigree recording''.)
All simulations used \fwdpp{} to implement a discrete-time Wright-Fisher population of $N$ diploid individuals,
simulated for $10N$ generations (details below).
Simulations without pedigree recording introduced neutral mutations at a rate
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected per-generation number of mutations per gamete 
and recombination breakpoints per diploid, respectively.
Simulations with pedigree recording introduced neutral mutations at the same rate
retrospectively, as described below, resulting in statistically identical simulation results.
% Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
% and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We ran simulations with different values of $N$ and varied the size of the genomic region according to the scaled recombination
parameter $\rho = 4Nr$.  
%% this was said already:
% When simulating neutral mutations, we kept the scaled mutation parameter, $\theta = 4N\mu$, equal to $\rho$.

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).  
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}). 
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

In our implementation, simulations with pedigree recording 
used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{
        \label{fig:runtimes_selection}
    Total run time per single simulation replicate as a function of region length.
    % measured as the scaled recombination parameter $\rho = 4Nr$. 
    Line color represents different diploid population sizes ($N$). 
    The left figure shows run times for standard simulations including neutral mutations. 
    The right column of panels shows run times of simulations that recorded the pedigree
    and added neutral mutations afterwards.
    The dashed line in the right panel shows results for an implementation using \fwdpy{} 
    where the pedigree simplification steps were handled in a separate thread of execution 
    and fitness calculations were parallelized.
    Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
    }
\end{figure}

\begin{figure}
    \includegraphics[]{sims/speedup}
    \caption{
        \label{fig:relative_speedup_selection}
        Relative speedup of simulations due to pedigree recording. 
        Each line shows the ratio of total run times of standard simulations to 
        those of simulations with pedigree recording. 
        Data points are taken from Figure~\ref{fig:runtimes_selection} 
        for simulations that ran to completion in both cases.
    }
\end{figure}

@@

%%%%%%
\subsection*{Tables for succinct tree sequences}

We now explain what we actually did to achieve a $50\times$ speedup.
The ``pedigree recording'' simulations above recorded information about each new individual
in a collection of tables that together define a \emph{succinct tree sequence}
(or, simply ``tree sequence'').
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants is recorded as \emph{sites} and \emph{mutations};
we give an example in Figure~\ref{fig:example_tree_sequence}. 
This formulation is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a more general class of tree topologies.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence.
Each node refers to a specific, distinct ancestor,
and so has a unique ``time'',
thought of as the node's birth time, which determines the height of any vertices
the node is associated with.
(Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.)
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born one and two units of time in the past, respectively.

@@

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees in this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding compact and algorithmically efficient.

Recovering the sequence of trees from this information is straightforward:
each point along the genome at which the tree topology changes
is accompanied by the end of some {edges} and the beginning of others.
Since each {edge} records the genomic interval
over which a given node inherits from a particular ancestor,
to construct the tree at a certain point in the genome
we need only retrieve all edges overlapping that point
and construct the corresponding tree.
To modify the tree to reflect the genealogy at a nearby location,
we simply remove those edges whose intervals do not overlap that location,
and add those new edges whose intervals do.
Incidentally, this property that edges naturally encode \emph{differences}
between nearby trees (e.g., as ``subtree prune and regraft'' moves)
allows for efficient algorithms to compute statistics of the genome sequence that take advantage
of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
records two things: its position on the genome and an ancestral state. 
For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'. 
If no mutations occur at a given site, all nodes inherit the ancestral state.

Each mutation records three things: the site at which it occurs,
the first node to inherit the mutation, and the derived state.
Thus, all nodes below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we used \msprime{} to simulate 500,000 samples of a
$200$ megabase chromosome with human-like parameters: $N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the node, edge, site and mutation tables (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB (giving a compression factor of 144,000 in
this instance).
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time just to read the information into memory.

@@

The facilities for working with succinct tree sequences are implemented as part
of the \msprime\ Python API, which provides a powerful platform for processing
tree topology and mutation data. The new portions of \msprime{} that we discuss
here are dedicated to tree sequence input and output using simple tables of
data, as described above, so we refer to this as the ``Tables API''.

The Tables API is primarily designed to facilitate efficient interchange of
data between programs or between different modules of the same program. We
adopted a `columnar' design, where all the values for a
particular column are stored in adjacent memory locations.
There are many advantages to columnar storage -- for example, since adjacent
values in memory are from the same column, they tend to compress well,
and suitable encodings can be chosen on a per-column basis~\citep{abadi2006integrating}.
A particular advantage of this approach is that it enables very
efficient copying of data, and in principle zero-copy data access
(where a data consumer reads directly from the memory of a producer).
Our implementation 
% uses the NumPy C API~\citep{walt2011numpy} to efficiently copy
efficiently copies data from Python as a NumPy array \citep{walt2011numpy} 
into the low-level C library used to manipulate tree sequences.
This architecture allows for data transfer rates of gigabytes per second
(impossible under any text-based approach), while retaining excellent portability.
NumPy's array interface provides a great deal of flexibility and efficiency,
and makes it straightforward to transfer data from sources
such as HDF5 \citep{hdf5} or Dask~\citep{dask}.
For small scale data and debugging purposes, a simple text based format is also supported.
